### 2015-02-05 src/3_TimeSeries.tex
Avance capitulo 2

\chapter{Time Series models}

@@

@@

There are several known features exhibited by financial instruments called
stylized facts, which have been empirically studied and some of them have been
documented only recently and accepted as truth. Stylized facts are usually
formulated in terms of qualitative properties of asset returns and may not be
precise enought to distinguish among different parametric models.
\cite{cont2001}.

@@

\begin{description}
\item[Dependence] autocorrelation in returns is largely insignificant. However,
autocorrelation in the absolute and squared returns is always positive and
significant and decays slowly. In addition, the autocorrelation in the absolute
returns is generally higher than the autocorrelation in the corresponding
squared returns.
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. Figure
\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
in average Monday has negative returns. However, this is changing in time,
figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
three months and it shows that Mondays and Wednesdays has positive returns in
average.

\begin{figure}[h!]
\centering
\begin{subfigure}[b]{0.45\textwidth}
 \includegraphics[width=\textwidth]{img/average_daily_change_1928_2007}
 \caption{Average daily return: 1928-2007}
 \label{fig:mondayeffect}
\end{subfigure}
\begin{subfigure}[b]{0.45\textwidth}
 \includegraphics[width=\textwidth]{img/thumb-average_daily_change_2007}
 \caption{Average daily return: 2007}
 \label{fig:mondayeffect2}
\end{subfigure}
\end{figure}

\end{description}

\begin{description}
 \item[Dependence]

 The returns
 autocorrelation function (ACF) of SPY is shown in figure
 \ref{fig:returnacf}.The figure shows how the correlations are significant even
 for very long lags, this imply that a long-memory process because of its
 long-range dependence. 
 Moreover, volatility tends to cluster, i.e. high volatility
 periods are generally followed by high volatility periods.

 \begin{figure}[h]
 \centering
 \includegraphics[width=0.9\textwidth]{plots/spy_returns_acf.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}

 \item[Returns distribution] The returns distribution has fat tails, i.e. the
 number of extreme events (either positive or negative) returns is larger than
 what is hypothesized by common data generation process (generally normal
 distribution assumption). In the markets, fat tails are an undesirable feature
 because of the additional risk they imply.  In the figure \ref{fig:returndist}
 is shown the SPY returns distribution based on daily dates from the period 1st
 July 1998 to 4th April 2013. The distribution was compared against the normal
 distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
 \item[Volume] refers to the level of trading activity in a market.
 \item[Calendar effects] days of week, weekend, holidays.
 \item[Intraday effects] time of the day.
 \item[Overnight effects]
\end{description}

\section{Volatility in the financial markets}

In financial markets, volatility is one of the key elements to model the
stochastic dynamic behaviour of financial assets. This is mainly because
volatility gives a measure of uncertainty about future returns and it is often
viewed as an indicator of the vulnerability of financial markets. It is also
important as an input parameter in problems like derivative pricing, hedging
and portfolio management.  For example, in option pricing we need to known
first the volatility of the underlying asset before pricing an option.

@@

\subsubsection{Types of volatility}

@@

\begin{description}
\item[realized volatility:] is also known as historic volatility and
it is the actual variance in the price of a stock over time.
Realized volatility is measured in terms of the standard deviation
using the historical stock prices. It is commonly calculated based on
intraday price returns:
\begin{equation}
\label{eq:retintra}
r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
\end{equation}
\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
intraday sample $n=2,\dots,N$. Realized volatility is defined as:
\begin{equation}
\label{eq:rv}
    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
\end{equation}
\noindent where $N$ is the number of intraday samples and $T$ is the
number of days. 
In order to include overnight returns, Hansen and
Lunde~\cite{hansen+lunde2005} introduced a scaling version of
realized volatility using the following definitions:
\begin{eqnarray}
r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
\end{eqnarray}
\noindent where equation~(\ref{eq: ret 1}) represents overnight 
returns and the volatility as 
equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
sample at day $t$. The scaled realized volatility $\rho(t)$ is 
defined as:
\begin{eqnarray}
\label{eq:srv}
\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
\end{eqnarray}
Realized volatility has also been defined as the absolute value return or as
the mean of the sum of intraday squared returns at short intervals of time. The
majority of research carried out in the literature obtain the daily volatility
as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
However, it has been proven that this measurement noise is too high for
observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen
and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could
result in an inferior model being chosen as the best one. The realized
volatility, as calculated by the cumulative sum of squared intraday returns and
shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an
inferior model.   
\item[implied volatility:] volatility not only can be extracted from returns
but it can also be derived from option or future pricing models.  The
volatility obtained corresponds to the market's prediction of future
volatility. In finance, an option is a derivative, that is, a contract which
gives the owner the right, but not the obligation to buy or sell an underlying
asset at a given price called strike price. An option can be executed at any
time before an expiration date previously defined no matter what price the
underlying asset has. For example, the Black-Scholes model~\cite{black1973}
determines the fair option value based on stock price, strike price, time to
option expiration, the interest rate and volatility. These are known or can be
easily obtained from the market, excepting by volatility which must be
estimated. However, rather than assuming a volatility a priori and computing
option prices from it, the model can be used to estimate volatility at given
prices, time to expiration and strike price. This obtained volatility is called
the implied volatility of an option. Additionally, some models obtain implied
volatility from futures (other derivative from prices). For instance, the
Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used
to determine future volatilities~\cite{hamidetal2004}. Higher implied
volatility is indicative of greater price fluctuation in either direction.
Implied volatility is found by determining the value which makes theoretical
prices equal to market prices. In this way volatility is ``implied'' by the
current market price of the stock.
\end{description}

@@

\subsubsection{Volatility methods}

@@

%Volatility models have been extensively
%studied~\cite{gatheral2006,poon+granger2003,knight2002} and some of  them
%include machine learning
%methods~\cite{hamidetal2004,donaldsonetal1997,shiyietal2008,shiyietal2010,gavrishchaka2006,vasilios2012}.  
%

\subsection{Stationary }
A strictly stationary times series $y_t$ is one for which the probabilistic behavior
of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
to that of the time shifted set, more precisely: \[ P\{y_{t_1} \leq
c_1,\dots,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,\dots,y_{t_L+h} \leq c_k\}
\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
$c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
assess it from a single data set. The weak version of this definition imposes
conditions only on the two first two moments.

\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$

Note that this definition includes the scalar case as time series of
vectors of dimension 1; in this scalar case we will write the time series in
nonbold format.

\subsection{Cointegration}
Cointegration concept was introduced by~\cite{engle1987} and implies that one or
more linear combinations of non-stationary variables are stationary even though
individually they are not.  Moreover ~\cite{stock+watson1988} observed that
cointegration reflects the common stochastic trends providing a useful way to
understand cointegration relationships. These common stochastic trends can be
also interpreted as a long-run equilibrium relationships.

The idea of cointegration was inmediatly adopted in finance since it could
represent their long-run relationship implied by economic
theory~\cite{laietAl1991}, \cite{lence+falk2005}.  Economic theory suggest that
economic time series are mean-reverting process and therefore, it reflects the
idea of that some set of variables cannot wander too far from each other.

On the other hand, the efficient markets hypothesis, also known as the random
walk theory states that current stock prices fully reflect available information
related to its value and there is no way to earn excess profits~\cite{fama1970}.
This means that if we have stock prices from a jointly efficient market, they
cannot be cointegrated \cite{granger1986}, \cite{dwyer1992}. However,
\cite{richards1995} claims that cointegration is directly at odds with market
efficiency, even though, there is no evidende that cointegration among asset
prices have implications about market efficiency~\cite{lence+falk2005}.

\subsection{Volatility}

\begin{description}%[leftmargin=0.4cm]\itemsep4pt
\item[realized volatility:] is also known as historic volatility and
it is the actual variance in the price of a stock over time.
Realized volatility is measured in terms of the standard deviation
using the historical stock prices. It is commonly calculated based on
intraday price returns:

\begin{equation}
\label{eq:retintra}
r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
\end{equation}

\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
intraday sample $n=2,\dots,N$. Realized volatility is defined as:

\begin{equation}
\label{eq:rv}
    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
\end{equation}

\noindent where $N$ is the number of intraday samples and $T$ is the
number of days.

In order to include overnight returns, Hansen and
Lunde~\cite{hansen+lunde2005} introduced a scaling version of
realized volatility using the following definitions:

\begin{eqnarray}
r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
\end{eqnarray}

\noindent where equation~(\ref{eq: ret 1}) represents overnight 
returns and the volatility as 
equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
sample at day $t$. The scaled realized volatility $\rho(t)$ is 
defined as:

\begin{eqnarray}
\label{eq:srv}
\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
\end{eqnarray}

Realized volatility has also been defined as the absolute value return or as the
mean of the sum of intraday squared returns at short intervals of time. The
majority of research carried out in the literature obtain the daily volatility
as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
However, it has been proven that this measurement noise is too high for
observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen and
Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could result
in an inferior model being chosen as the best one. The realized volatility, as
calculated by the cumulative sum of squared intraday returns and shown in
equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an inferior
model.

\item[implied volatility:] volatility not only can be extracted from returns but
it can also be derived from option or future pricing models.  The volatility
obtained corresponds to the market's prediction of future volatility. In
finance, an option is a derivative, that is, a contract which gives the owner
the right, but not the obligation to buy or sell an underlying asset at a given
price called strike price. An option can be executed at any time before an
expiration date previously defined no matter what price the underlying asset
has. For example, the Black-Scholes model~\cite{black1973} determines the fair
option value based on stock price, strike price, time to option expiration, the
interest rate and volatility. These are known or can be easily obtained from the
market, excepting by volatility which must be estimated. However, rather than
assuming a volatility a priori and computing option prices from it, the model
can be used to estimate volatility at given prices, time to expiration and
strike price. This obtained volatility is called the implied volatility of an
option. Additionally, some models obtain implied volatility from futures (other
derivative from prices). For instance, the Barone-Adesi and Whaley futures
option model~\cite{baroneetal1987} is also used to determine future
volatilities~\cite{hamidetal2004}. Higher implied volatility is indicative of
greater price fluctuation in either direction. Implied volatility is found by
determining the value which makes theoretical prices equal to market prices. In
this way volatility is ``implied'' by the current market price of the stock.

\end{description}

\subsubsection{Volatility methods}

%\section{Black-Scholes formula}
%
%In finance, an option is a derivative, that is, a contract which gives the owner
%the right, but not the obligation to buy or sell an underlying asset at a given
%price called strike price. An option can be executed at any time before an
%expiration date previously defined no matter what price the underlying asset
%has. 
%
%The Black-Scholes formula~\cite{black1973}, developed in the early 1970's, Myron
%Scholes, Robert Merton and Fisher Black,  allows to determine an option value
%$V$ based on the underlying asset price $S(t)$ at a time $t$ and the following
%constant parameters: 
%
%\begin{description}
%\item [$\sigma$:] underlying asset price volatility which measures the standard
%deviation of the returns
%\item [$\mu$:] underlying asset drift which is a measure of the average rate of
%growth of the stock
%\item[$E$:] option strike or excersice price
%\item[$T$:] option date of expiry
%\item[$t$:] current time
%\item[$r$:] risk-free interest rate
%\end{description}
%
%\subsection{Stock price model}
%
%The Black-Scholes model assume that the underlying price $S$ follows a lognormal random walk:
%
%\begin{equation}\label{eq:stockprice}
%dS = \mu S dt + \sigma S dB
%\end{equation}
%
%\noindent where $B$ is a Brownian motion. This stochastic differential equation
%has two components: a deterministic term given by $\mu S dt$ and a random term
%given by $\sigma S dX$.
%
%\subsection{Brownian motion}
%
%A brownian motion $B$ (also called a Wiener process) is a stochastic process
%characterized by the three following properties:
%
%\begin{description}
%\item[Continuity:] $B(t)$ is a continuos function
%\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$
%and variance $t-s$.
%\item[Independence of increments:] for every choice of nonnegative real numbers
%$0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random
%variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly
%independent.
%\end{description}
%
%\subsection{Stochastic differential equation}
%
%An stochastic differential integral has the form:
%
%
%\begin{equation}
%W(T)=\int_0^T f(t)dB(t) \, .
%\end{equation}
%
%\noindent This equations is also expressed in an abbreviate form:
%
%\begin{equation}
%dW = f(t) dB \, .
%\end{equation}
%
%\noindent Therefore, the integral form of the stock price model shown in
%equation (\ref{eq:stockprice}) is:
%
%\begin{equation}
%S(T)=\int_0^T \mu S(t) dt + \int_0^T \sigma S(t) dB(t)
%\end{equation}
%
%\subsection{Ito's Lemma}
%
%Ito's lemma is used to find the differential of a time dependent function of a
%stochastic process. In option pricing we need to find the option price $V(S(t))$
%which depends on a stochastic stock price model $S(t)$.  $V(S,t)$ is required to
%be  differentiable function of $S$ and once differentiable function of $t$.
%
%
%
%
%
%%These are known or can be easily obtained from the market, excepting by
%volatility which must be estimated. However, rather than assuming a volatility
%a priori and computing option prices from it, the model can be used to estimate
%volatility at given prices, time to expiration and strike price. This obtained
%volatility is called the implied volatility of an option. Additionally, some
%models obtain implied volatility from futures (other derivative from prices). 
%
%
% 
%\section{Realized Volatility Models}
%
%\section{GARCH} test \section{ARFIMA} \section{Stochastic Volatility}
%
%\section{Volatility definition}
%
%\begin{itemize}
%
%\item Is the size of the price movement.
%
%\item Variance is a measure of distribution of returns and is not neccesarily
%bound by any time period.  Volatility is a measure of the standard deviation
%(square root of the variance) over a certain time interval. In finance,
%variance and volatility both gives you a sense of an asset's risk. Variance
%gives you a sense of the risk in the asset over its lifetime, while volatility
%gives you a sense of the movement of the asset in eg. the past month or the
%past year.
%
%
%\item The main underlying difference is in their definition. Variance has a
%fixed mathematical definition, however volatility does not as such. Volatility
%is said to be the measure of fluctuations of a process.
%
%Volatility is a subjective term, whereas variance is an objective term i.e.
%given the data you can definitely find the variance, while you can't find
%volatility just having the data. Volatility is associated with the process, and
%not with the data.
%
%In order to know the volatility you need to have an idea of the process i.e you
%need to have an observation of the dispersion of the process. All the different
%processes will have different methods to compute volatilities based on the
%underlying assumptions of the process.  \end{itemize}
%

\subsection{Vector Autorregresive Models}\label{sec:varvec}

%VECM is a special case of VAR model and both describe the joint behaviour
%of a set of variables.

The VAR($p$) model is a general framework describing the behaviour of a
set of $l$ endogenous variables as a linear combination of their last $p$
values, where $l,p\in\mathbb{N}$. 
In our case, each one of these $l$ variables is a scalar time series
$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
at time $t$ by the the vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

@@

VECM is a special form of a VAR model for I(1) variables that are also
cointegrated~\cite{banerjee1993}.

@@

>>>>

\chapter{Financial Time Series}

@@

\subsection{Stationary }
A strictly stationary times series $Y_t$ is one for which the probabilistic behaviour
of every collection of values $\{Y_{t_1},Y_{t_2},\dots,Y_{t_L}\}$ is identical
to that of the time shifted set, more precisely: \[ P\{Y_{t_1} \leq
c_1,\dots,Y_{t_L} \leq c_L\} = P\{Y_{t_1+h} \leq c_1,\dots,Y_{t_L+h} \leq c_L\}
\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
$c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
assess it from a single data set. The weak version of this definition imposes
conditions only on the two first two moments.

\noindent with $\lambda(s,t) = E[(Y_s-\mu)(Y_t - \mu)]$

In finance, a shock represents an unexpected change in a variable or a in its error term in a particular time period. For stationary time series, shocks to the system will gradually die away. That is, a shock during time $t$ will have a smaller effect in time $t+1$, a smaller effect on $t+2$ and so on. For non-stationary data, the persistence of shocks will always be infinite. The use of non-stationary data can lead to spurious regressions. If two stationary variables are generated as independent random series and are trending over time, when one of those variables is regressed on the other, they could have a high $R^2$ even if the two are totally unrelated. So, if standard regression techniques are applied to non-stationary data, it could look good under standard measures but valueless \cite{brooks2002}.

Note that this definition includes the scalar case as time series of
vectors of dimension 1; in this scalar case we will write the time series in
non-bold format.

\subsection{Cointegration}
Cointegration concept was introduced by Engle in 1987 \cite{engle1987} and implies that one or
more linear combinations of non-stationary variables are stationary even though
individually they are not.

Stock and Watson in 1988 \cite{stock+watson1988} observed that
cointegration reflects the common stochastic trends providing a useful way to
understand cointegration relationships. These common stochastic trends can be
also interpreted as a long-run equilibrium relationships.

The idea of cointegration was immediately adopted in finance since it could
represent their long-run relationship implied by economic
theory~\cite{laietAl1991}, \cite{lence+falk2005}.  Economic theory suggest that
economic time series are mean-reverting process and therefore, it reflects the
idea of that some set of variables cannot wander too far from each other.

On the other hand, the efficient markets hypothesis, also known as the random
walk theory states that current stock prices fully reflect available information
related to its value and there is no way to earn excess profits~\cite{fama1970}.
This means that if we have stock prices from a jointly efficient market, they
cannot be cointegrated \cite{granger1986}, \cite{dwyer1992}. However,
\cite{richards1995} claims that cointegration is directly at odds with market
efficiency, even though, there is no evidence that cointegration among asset
prices have implications about market efficiency~\cite{lence+falk2005}.

@@

There are several known features exhibited by financial instruments called
stylized facts, which have been empirically studied and some of them have been
documented only recently and accepted as truth. Stylized facts are usually
formulated in terms of qualitative properties of asset returns and may not be
precise enough to distinguish among different parametric models.
\cite{cont2001}.

@@

\begin{description}
\item[Dependence] autocorrelation in returns is largely insignificant. However,
autocorrelation in the absolute and squared returns is always positive and
significant and decays slowly. In addition, the autocorrelation in the absolute
returns is generally higher than the autocorrelation in the corresponding
squared returns. The returns autocorrelation function (ACF) of SPY is shown in figure
 \ref{fig:returnacf}. The figure shows how the correlations are significant even
 for very long lags, this imply that a long-memory process because of its
 long-range dependence. 
 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
The returns distribution tails are larger than
 what is hypothesised by common data generation process (generally normal
 distribution assumption). In the markets, fat tails are an undesirable feature
 because of the additional risk they imply.  In the figure \ref{fig:returndist}
 is shown the SPY returns distribution based on daily dates from the period 1st
 July 1998 to 4th April 2013. The distribution was compared against the normal
 distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. Figure
\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
in average Monday has negative returns. However, this is changing in time,
figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
three months and it shows that Mondays and Wednesdays has positive returns in
average.

\begin{figure}[h!]
\centering
\begin{subfigure}[b]{0.45\textwidth}
 \includegraphics[width=\textwidth]{img/average_daily_change_1928_2007}
 \caption{Average daily return: 1928-2007}
 \label{fig:mondayeffect}
\end{subfigure}
\begin{subfigure}[b]{0.45\textwidth}
 \includegraphics[width=\textwidth]{img/thumb-average_daily_change_2007}
 \caption{Average daily return: 2007}
 \label{fig:mondayeffect2}
\end{subfigure}
\end{figure}
\end{description}

\section{Volatility in the financial markets}
In financial markets, volatility is one of the key elements to model the
stochastic dynamic behaviour of financial assets. This is mainly because
volatility gives a measure of uncertainty about future returns and it is often
viewed as an indicator of the vulnerability of financial markets. It is also
important as an input parameter in problems like derivative pricing, hedging
and portfolio management.  For example, in option pricing we need to known
first the volatility of the underlying asset before pricing an option.

@@

Despite the fact variance and volatility are related, they are not the same concept. Variance is a measure of distribution of returns and is not necessarily
bound by any time period.  Volatility is a measure of the standard deviation
(square root of the variance) over a certain time interval. In finance,
variance and volatility both gives you a sense of an asset's risk. Variance
gives you a sense of the risk in the asset over its lifetime, while volatility
gives you a sense of the movement of the asset in, for example, the past month or the
past year. The main underlying difference is in their definition. Variance has a
fixed mathematical definition, however volatility does not as such. Volatility
is said to be the measure of fluctuations of a process.

Volatility is a subjective term, whereas variance is an objective term i.e.
given the data you can definitely find the variance, while you can't find
volatility just having the data. Volatility is associated with the process, and
not with the data. In order to know the volatility you need to have an idea of the process i.e you
need to have an observation of the dispersion of the process. All the different
processes will have different methods to compute volatilities based on the
underlying assumptions of the process.

\subsection{Types of volatility}

@@

\subsubsection{realized volatility} is also known as historic volatility and
it is the actual variance in the price of a stock over time.
Realized volatility is measured in terms of the standard deviation
using the historical stock prices. It is commonly calculated based on
intraday price returns:
\begin{equation}
\label{eq:retintra}
r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
\end{equation}
\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
intraday sample $n=2,\dots,N$. Realized volatility is defined as:
\begin{equation}
\label{eq:rv}
    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
\end{equation}
\noindent where $N$ is the number of intraday samples and $T$ is the
number of days. 
In order to include overnight returns, Hansen and
Lunde~\cite{hansen+lunde2005} introduced a scaling version of
realized volatility using the following definitions:
\begin{eqnarray}
r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
\end{eqnarray}
\noindent where equation~(\ref{eq: ret 1}) represents overnight 
returns and the volatility as 
equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
sample at day $t$. The scaled realized volatility $\rho(t)$ is 
defined as:
\begin{eqnarray}
\label{eq:srv}
\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
\end{eqnarray}
Realized volatility has also been defined as the absolute value return or as
the mean of the sum of intraday squared returns at short intervals of time. The
majority of research carried out in the literature obtain the daily volatility
as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
However, it has been proven that this measurement noise is too high for
observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen
and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could
result in an inferior model being chosen as the best one. The realized
volatility, as calculated by the cumulative sum of squared intraday returns and
shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an
inferior model.   
\subsubsection{implied volatility volatility} not only can be extracted from returns
but it can also be derived from option or future pricing models.  The
volatility obtained corresponds to the market's prediction of future
volatility. In finance, an option is a derivative, that is, a contract which
gives the owner the right, but not the obligation to buy or sell an underlying
asset at a given price called strike price. An option can be executed at any
time before an expiration date previously defined no matter what price the
underlying asset has. For example, the Black-Scholes model~\cite{black1973}
determines the fair option value based on stock price, strike price, time to
option expiration, the interest rate and volatility. These are known or can be
easily obtained from the market, excepting by volatility which must be
estimated. However, rather than assuming a volatility a priori and computing
option prices from it, the model can be used to estimate volatility at given
prices, time to expiration and strike price. This obtained volatility is called
the implied volatility of an option. Additionally, some models obtain implied
volatility from futures (other derivative from prices). For instance, the
Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used
to determine future volatilities~\cite{hamidetal2004}. Higher implied
volatility is indicative of greater price fluctuation in either direction.
Implied volatility is found by determining the value which makes theoretical
prices equal to market prices. In this way volatility is ``implied'' by the
current market price of the stock.

In finance, an option is a derivative, that is, a contract which gives the owner
the right, but not the obligation to buy or sell an underlying asset at a given
price called strike price. An option can be executed at any time before an
expiration date previously defined no matter what price the underlying asset
has.

The Black-Scholes formula~\cite{black1973}, developed in the early 1970's, Myron
Scholes, Robert Merton and Fisher Black,  allows to determine an option value
$V$ based on the underlying asset price $S(t)$ at a time $t$ and the following
constant parameters:

\begin{description}
\item [$\sigma$:] underlying asset price volatility which measures the standard
deviation of the returns
\item [$\mu$:] underlying asset drift which is a measure of the average rate of
growth of the stock
\item[$E$:] option strike or excersice price
\item[$T$:] option date of expiry
\item[$t$:] current time
\item[$r$:] risk-free interest rate
\end{description}

The Black-Scholes model assume that the underlying price $S$ follows a lognormal random walk:

\begin{equation}\label{eq:stockprice}
dS = \mu S dt + \sigma S dB
\end{equation}

\noindent where $B$ is a Brownian motion. This stochastic differential equation
has two components: a deterministic term given by $\mu S dt$ and a random term
given by $\sigma S dX$.

A brownian motion $B$ (also called a Wiener process) is a stochastic process
characterized by the three following properties:

\begin{description}
\item[Continuity:] $B(t)$ is a continuos function
\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$
and variance $t-s$.
\item[Independence of increments:] for every choice of nonnegative real numbers
$0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random
variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly
independent.
\end{description}

An stochastic differential integral has the form:

\begin{equation}
W(T)=\int_0^T f(t)dB(t) \, .
\end{equation}

\noindent This equations is also expressed in an abbreviate form:

\begin{equation}
dW = f(t) dB \, .
\end{equation}

\noindent Therefore, the integral form of the stock price model shown in
equation (\ref{eq:stockprice}) is:

\begin{equation}
S(T)=\int_0^T \mu S(t) dt + \int_0^T \sigma S(t) dB(t)
\end{equation}

Ito's lemma is used to find the differential of a time dependent function of a
stochastic process. In option pricing we need to find the option price $V(S(t))$
which depends on a stochastic stock price model $S(t)$.  $V(S,t)$ is required to
be  differentiable function of $S$ and once differentiable function of $t$.

These are known or can be easily obtained from the market, excepting by
volatility which must be estimated. However, rather than assuming a volatility
a priori and computing option prices from it, the model can be used to estimate
volatility at given prices, time to expiration and strike price. This obtained
volatility is called the implied volatility of an option. Additionally, some
models obtain implied volatility from futures (other derivative from prices).

@@

\subsection{Volatility methods}

@@

\section{Vector Autoregressive Models}\label{sec:varvec}

The vector autoregressive VAR($p$) model is one of the most easy to use, successful and flexible models for the analysis of multivariate time series. It is a natural extension of the univariate autoregressive (AR) model. The VAR model has proven to be useful for describing the dynamic behaviour of economic and financial time series.
VAR is a general framework describing the behaviour of a
set of $l$ endogenous variables as a linear combination of their last $p$
values, where $l,p\in\mathbb{N}$. 
In our case, each one of these $l$ variables is a scalar time series
$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
at time $t$ by the the vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

@@

\section{Vector error correction model}
Vector error correction model (VECM) is a special form of a VAR model for I(1) variables that are also
cointegrated~\cite{banerjee1993}.

@@

\section{Unit root tests}

\section{Johansen method}
